\section{Hybrid Assembly}
\begin{note}
Like the previous examples, the data you will examine in this exercise is again
from Staphylococcus aureus which has a genome of around 3MB. The reads are 454
single end and Illumina paired end with an insert size of 170 bp.
You already downloaded the required reads from the SRA in previous exercises.
Specifically, the run data (SRR022863, SRR000892, SRR000893) from the SRA
experiments SRX007709 and SRX000181.
\end{note}

\begin{information}
The following exercise focuses on handing 454 long reads and paired-end reads
with velvet and the differences in setting parameters.
\end{information}

\begin{steps}
First move to the directory you made for this exercise, make a suitable named
directory for the exercise and check if all the three files are in place:
\begin{lstlisting}
cd ~/NGS/velvet/part3
mkdir SRR000892-SRR022863 
cd SRR000892-SRR022863
# 454 single end data
ln -s ~/NGS/Data/SRR00089[2-3].fastq.gz ./
# illumina paired end data
ln -s ~/NGS/Data/SRR022863_?.fastq.gz ./
\end{lstlisting}
\end{steps}

\begin{warning}
The following command will run for a LONG time. This indicated the amount of
calculations being preformed by Velvet to reach a conclusion. To wait for velvet
to finish would exceed the time available in this workshop, but it is up to you
to either let it run over night or kill the process by using the key combination
\texttt{CTRL+c}.
\begin{lstlisting}
velveth run_25 25 -fmtAuto -create_binary -long SRR00089?.fastq.gz -shortPaired -separate SRR022863_1.fastq.gz SRR022863_2.fastq.gz
time velvetg run_25
\end{lstlisting}

\begin{steps}
If you have decided to continue, we already inspected the weighted histograms
for the short and long read library separately, you can reuse this for the
cut-off values:
\begin{lstlisting}
time velvetg run_25 -cov_cutoff 7 -long_cov_cutoff 9
\end{lstlisting}
\end{steps}

\begin{questions}
What are your conclusions using Velvet in an hybrid assembly?
\begin{answer}
17 min:  time velvetg run\_25
\end{answer}
\end{questions}

\end{warning}