% Define the top matter
\setModuleTitle{\textit{de novo} Genome Assembly}
\setModuleAuthors{%
  Matthias Haimel \mailto{mhaimel@ebi.ac.uk}\\
  Nathan S. Watson-Haigh \mailto{nathan.watson-haigh@awri.com.au}
}
\setModuleContributions{%
  
}

%  Start: Module Title Page
\chapter{\moduleTitle}
\newpage
% End: Module Title Page

\section{Key Learning Outcomes}

After completing this practical the trainee should be able to:
\begin{itemize}
  \item Compile velvet with appropriate compile-time parameters set for a specific analysis
  \item Be able to choose appropriate assembly parameters
  \item Assemble a set of single-ended reads
  \item Assemble a set of paired-end reads from a single insert-size library
%  \item Assemble a set of long 454 single-end reads
%  \item Assemble multiple paired-end libraries with different insert-sizes
%  \item Assemble a hybrid set of long 454 single-end reads and a set of paired-end reads
  \item Be able to visualise an assembly in AMOS Hawkeye
  \item Understand the importance of using paired-end libraries in \textit{de
        novo} genome assembly
\end{itemize}

\section{Resources You'll be Using}
Although we have provided you with an environment which contains all the tools
and data you will be using in this module, you may like to know where we have
sourced those tools and data from.
 
\subsection{Tools Used}
\begin{description}[style=multiline,labelindent=0cm,align=left,leftmargin=1cm]
  \item[Velvet] \hfill\\
  	\url{http://www.ebi.ac.uk/~zerbino/velvet/}
  \item[AMOS Hawkeye] \hfill\\
    \url{http://apps.sourceforge.net/mediawiki/amos/index.php?title=Hawkeye}
  \item[gnx-tools] \hfill\\
    \url{https://github.com/mh11/gnx-tools}
  \item[FastQC] \hfill\\
    \url{http://www.bioinformatics.bbsrc.ac.uk/projects/fastqc/}
  \item[R] \hfill\\
    \url{http://www.r-project.org/}
\end{description}

\subsection{Sources of Data}
\begin{itemize}
\item \url{ftp://ftp.ensemblgenomes.org/pub/release-8/bacteria/fasta/Staphylococcus/s_aureus_mrsa252/dna/s_aureus_mrsa252.EB1_s_aureus_mrsa252.dna.chromosome.Chromosome.fa.gz}
% single end data
\item \url{http://www.ebi.ac.uk/ena/data/view/SRS004748}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022825/SRR022825.fastq.gz}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022823/SRR022823.fastq.gz}
% paired-end data
%\item \url{http://www.ebi.ac.uk/ena/data/view/SRS004748}
\item \url{http://www.ebi.ac.uk/ena/data/view/SRX008042}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022852/SRR022852_1.fastq.gz}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022852/SRR022852_2.fastq.gz}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR023/SRR023408/SRR023408_1.fastq.gz}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR023/SRR023408/SRR023408_2.fastq.gz}
% 454 Module data
%\item \url{http://www.ebi.ac.uk/ena/data/view/SRX000181}
%\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR000/SRR000892/SRR000892.fastq.gz}
%\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR000/SRR000893/SRR000893.fastq.gz}
% Multiple insert sizes data
%\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022852/SRR022852_1.fastq.gz}
%\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022852/SRR022852_2.fastq.gz}
% hybrid data
% 454
\item \url{http://www.ebi.ac.uk/ena/data/view/SRX000181}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR000/SRR000892/SRR000892.fastq.gz}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR000/SRR000893/SRR000893.fastq.gz}
% illumina paired-end
\item \url{http://www.ebi.ac.uk/ena/data/view/SRX007709}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022863/SRR022863_1.fastq.gz}
\item \url{ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR022/SRR022863/SRR022863_2.fastq.gz}
\end{itemize}

\newpage

\section{Introduction}
The aim of this module is to become familiar with performing \textit{de novo}
genome assembly using Velvet, a de Bruijn graph based assembler, on a variety of
sequence data.

\section{Prepare the Environment}
\begin{information}
The first exercise should get you a little more comfortable with the computer
environment and the command line.
\end{information}

\begin{steps}
First make sure that you are in your home directory by typing:
\begin{lstlisting}
cd
\end{lstlisting}

and making absolutely sure you're there by typing:
\begin{lstlisting}
pwd
\end{lstlisting}

Now create sub-directories for this and the two other velvet practicals. All
these directories will be made as sub-directories of a directory for the whole
course called NGS. For this you can use the following commands:
\begin{lstlisting}
mkdir -p NGS/velvet/{part1,part2,part3}
\end{lstlisting}
\end{steps}

\begin{information}
The \texttt{-p} tells \texttt{mkdir} (make directory) to make any parent
directories if they don't already exist. You could have created the above directories one-at-a-time
by doing this instead:
\begin{lstlisting}
mkdir NGS
mkdir NGS/velvet
mkdir NGS/velvet/part1
mkdir NGS/velvet/part2
mkdir NGS/velvet/part3
\end{lstlisting}
\end{information}

\begin{steps}
After creating the directories, examine the structure and move into the
directory ready for the first velvet exercise by typing:
\begin{lstlisting}
ls -R NGS
cd NGS/velvet/part1
pwd
\end{lstlisting}

\end{steps}

\section{Downloading and Compiling Velvet}
\begin{note}
For the duration of this workshop, all the software you require has been set up
for you already. This might not be the case when you return to ``real life''. Many
of the programs you will need, including velvet, are quite easy to set up, it
might be instructive to try a couple.
\end{note}

\begin{information}
Although you will be using the preinstalled version of velvet, it is useful to
know how to compile velvet as some of the parameters you might like to control
can only be set at compile time. You can find the latest version of velvet at:

{\centering
\url{http://www.ebi.ac.uk/~zerbino/velvet/}
 
}

You could go to this URL and download the latest velvet version, or
equivalently, you could type the following, which will download, unpack,
inspect, compile and execute your locally compiled version of velvet:
\begin{lstlisting}
cd ~/NGS/velvet/part1
pwd
tar xzf ~/NGS/Data/velvet_1.2.07.tgz
ls -R
cd velvet_1.2.07
make
./velveth
\end{lstlisting}
The standout displayed to screen when 'make' runs may contain an error message but it is ignored
\end{information}

\begin{steps}
Take a look at the executables you have created. They will be displayed as green
by the command:
\begin{lstlisting}
ls --color=always
\end{lstlisting}
\end{steps}

\begin{note}
The switch \texttt{--color}, instructs that files be coloured according to their
type. This is often the default but we are just being explicit. By specifying
the value \texttt{always}, we ensure that colouring is always applied, even from
a script.
\end{note}

\begin{steps}
Have a look of the output the command produces and you will see that
\texttt{MAXKMERLENGTH=31} and \texttt{CATEGORIES=2} parameters were passed into
the compiler.

This indicates that the default compilation was set for de Bruijn graph k-mers of
maximum size 31 and to allow a maximum of just 2 read categories. You can
override these, and other, default configuration choices using command line
parameters. Assume, you want to run velvet with a k-mer length of 41 using 3
categories, velvet needs to be recompiled to enable this functionality by
typing:
\begin{lstlisting}
make clean
make MAXKMERLENGTH=41 CATEGORIES=3
./velveth
\end{lstlisting}
\end{steps}

\begin{questions}
Discuss with the persons next to you the following questions:\\
What are the consequences of the parameters you have given make for velvet?
\begin{answer}
MAXKMERLENGTH: increase the max k-mer length from 31 to 41

CATEGORIES: paired-end data require to be put into separate categories. By
increasing this parameter from 2 to 3 allows you to process 3 paired / mate-pair
libraries and unpaired data.
\end{answer}

Why does Velvet use k-mer 31 and 2 categories as default?
\begin{answer}
Possibly a number of reason:\\
  - odd number to avoid palindromes\\
  - The first reads were very short (20-40 bp) and there were hardly any paired-end data\\
     around so there was no need to allow for longer k-mer lengths / more categories.\\
  - For programmers: 31 bp get stored in 64 bits (using 2bit encoding)
\end{answer}

Should you get better results by using a longer k-mer length?
\begin{answer}
If you can achieve a good k-mer coverage - yes.
\end{answer}
\end{questions}

\begin{bonus}
velvet can also be used to process SOLID colour space data. To do this you need
a further make parameter. With the following command clean away your last
compilation and try the following parameters:
\begin{lstlisting}
make clean
make MAXKMERLENGTH=41 CATEGORIES=3 color
./velveth_de
\end{lstlisting}
\end{bonus}

\begin{questions}
What effect would the following compile-time parameters have on velvet:\\
\texttt{OPENMP=Y}
\begin{answer}
Turn on multithreading
\end{answer}

\texttt{LONGSEQUENCES=Y}
\begin{answer}
Assembling reads / contigs longer than 32kb long
\end{answer}

\texttt{BIGASSEMBLY=Y}
\begin{answer}
Using more than 2.2 billion reads
\end{answer}

\texttt{VBIGASSEMBLY=Y}
\begin{answer}
Not documented yet
\end{answer}

\texttt{SINGLE\_COV\_CAT=Y}
\begin{answer}
Merge all coverage statistics into a single variable - save memory
\end{answer}

\end{questions}

\begin{note}
For a further description of velvet compile and runtime parameters please see
the velvet Manual: \url{https://github.com/dzerbino/velvet/wiki/Manual}
\end{note}

\newpage
\input{de_novo/velvet/single_end.tex}

\input{de_novo/velvet/paired_end.tex}

%\input{de_novo/velvet/454_reads.tex}

%\input{de_novo/velvet/multiple_insert_sizes.tex}

\begin{advanced}
\input{de_novo/velvet/hybrid.tex}
\end{advanced}

